---
title: "Project 1"
author: "Jason Ballinger, Jason Perez, Megan Juza"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---
```{r setup, include = FALSE}

#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1
```{r ob1}
# Objective 1: Determine where COVID-19 originated from

library(readr)

# Import data sets
cases <- data.frame(read_delim(file="time_series_covid19_confirmed_global.csv", delim=","))
deaths <- data.frame(read_delim(file="time_series_covid19_deaths_global.csv", delim=","))

# Highest confirmed cases & deaths for first day (1/22/2020)
highestCasesIndex <- which.max(cases$X1.22.20)
highestDeathsIndex <- which.max(cases$X1.22.20)

# Get location given the index
highestCasesLocation <- paste(cases$Province.State[highestCasesIndex], cases$Country.Region[highestCasesIndex])
highestDeathsLocation <- paste(deaths$Province.State[highestDeathsIndex], cases$Country.Region[highestDeathsIndex])

# Check if the locations are the same
if (highestCasesLocation == highestDeathsLocation) {
  print(paste("COVID-19 originated from:", highestCasesLocation))
} else {
  print("Unable to find COVID-19's origin location")
}
```

### Objective 2
```{r ob2}
# Objective 2: Where is the most recent area to have a first confirmed case?

library(readr)

# Import data set
cases <- data.frame(read_delim(file="time_series_covid19_confirmed_global.csv", delim=","))

# Find the indices of the most recent first case
recentRow <- NA
recentCol <- NA

for (i in 1:nrow(cases)) {
  row <- cases[i, ] # Splice the row (state/country)
  for (j in 5:ncol(cases)) {
    # Check for first case
    if (cases[i, j] >= 1) {
      # Check if it has a more recent first case
      if (j > recentCol || is.na(recentCol)) { 
        recentRow <- i
        recentCol <- j
      }
      break
    }
  }
}

# return the state + country value for the highest index
print(paste("Region with the newest first case is:", cases[recentRow, 1], cases[recentRow, 2]))
```

### Objective 3
```{r ob3}

```

### Objective 4
```{r ob4}

```

### Objective 5
```{r ob5}

```

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```




